Visualizing Project Evolution Through Abstract Syntax Tree Analysis

### 2016-04-10 paper/paper.tex
Fixed some formatting.

In the following sections of the paper we will introduce a tool, TypeV, that we developed that visualizes changes to types used in a project over time. TypeV is able to extract type information from a repository by computing commit-by-commit abstract syntax tree differencing, which we describe in section~\ref{sec:methodology}. We seek to answer the following questions:
\\ \\
\textbf{RQ1:} How can we extract type and method usage over time from repositories? \\ \\
\textbf{RQ2:} Does the type information give any interesting or useful insight into the project? \\ \\
\textbf{RQ3:} How does type coverage compare to file coverage?

@@

The analyzed data would be difficult to analyze if displayed as its raw text form; instead, we created an HTML5 web visualization application for intended for developers, project managers, and researchers to use~\ref{fig:main-view}.

TypeV provide a variety of views to help organize the data and provide a finer level of detail. The main view displays a series of stacked bar graphs which show changes to types over a period of time in a project~\ref{fig:main-view}. The blue and orange portions account for the number of additions and deletions in a similar fashion to GitHub’s code frequency graph~\ref{fig:GH_code_freq}. The height of each bar shows the number of changes to a type in a smaller unit of time. This is calculated by multiplying the bar height by the number of changes during that time period divided by the maximum number of changes attested for this type. A bar which fills the row vertically is the period with the largest number of changes to that type out of the visible time periods. The other bars in the row are proportional to that maximum number. The width of each bar represents a set time period and can be changed to an hour, day, week, or month. The rows are sorted so that the graph for the most used types appear the top and are in descending order.

When a user mouses over a specific bar in the graph they are given simple statistics for the type and time period at that location. This includes the fully-qualified name of the type, the number of commits, the number of authors, the number of additions, the number of deletions, and the start and end date of that time period. If the user clicks on the bar, they are given more detailed statistics which list the authors who made the commits and each commit message~\ref{fig:typev-deets}.

@@

In the following sections, we present {N} case studies demonstrating the extra insight that the TypeV method makes available.

@@

Apache Bookkeeper deprecated the method AbstractConfiguration.getLedgerManagerType() in release 4.2.0~\footnote{\url{https://bookkeeper.apache.org/docs/r4.2.0/apidocs/deprecated-list.html}} and replaced it with AbstractConfiguration.getLedgerManagerFactoryClass(). If we look at getLedgerManagerType() we can see at the time of the start of the deprecation all instances were removed and getLedgerManagerFactoryClass() was added. After the deprecation, we see no more changes to getLedgerManagerType() and only some additions of getLedgerManagerFactoryClass().

@@

To determine if type coverage as returned by TypeV gives different information than file coverage alone, we computed the following. For four projects (Antlr, Apache Bookkeeper, Apache Curator, Apache Tika), we counted the file coverage (cumulative amount of files touched per author), as well as type coverage (cumulative amount of types touched per author) on a commit-by-commit basis. That is, each datum is represents a commit with two dimensions: file coverage and type coverage. Data is sorted in ascending order of commit date. Table~\ref{tab:summary} lists a summary of the data

@@

@@

Authors can pull other projects into the repository and these additions will count towards the author's changes.

We do not look at other branches besides the master; therefore, we might miss some of the work done by an author.

>>>>

In the following sections of the paper we will introduce a tool, TypeV, that we developed that visualizes changes to types used in a project over time. TypeV is able to extract type information from a repository by computing commit-by-commit abstract syntax tree differencing, which we describe in section~\ref{sec:methodology}. We seek to answer the following questions: \\

\textbf{RQ1:} How can we extract type and method usage over time from repositories? \\ \\
\textbf{RQ2:} Does the type information give any interesting or useful insight into the project? \\ \\
\textbf{RQ3:} How does type coverage compare to file coverage?

@@

The analyzed data would be difficult to analyze if displayed as its raw text form; instead, we created an HTML5 web visualization application for intended for developers, project managers, and researchers to use (Figure~\ref{fig:main-view}).

TypeV provide a variety of views to help organize the data and provide a finer level of detail. The main view displays a series of stacked bar graphs which show changes to types over a period of time in a project (Figure~\ref{fig:main-view}). The blue and orange portions account for the number of additions and deletions in a similar fashion to GitHub’s code frequency graph (Figure~\ref{fig:GH_code_freq}). The height of each bar shows the number of changes to a type in a smaller unit of time. This is calculated by multiplying the bar height by the number of changes during that time period divided by the maximum number of changes attested for this type. A bar which fills the row vertically is the period with the largest number of changes to that type out of the visible time periods. The other bars in the row are proportional to that maximum number. The width of each bar represents a set time period and can be changed to an hour, day, week, or month. The rows are sorted so that the graph for the most used types appear the top and are in descending order.

When a user mouses over a specific bar in the graph they are given simple statistics for the type and time period at that location. This includes the fully-qualified name of the type, the number of commits, the number of authors, the number of additions, the number of deletions, and the start and end date of that time period. If the user clicks on the bar, they are given more detailed statistics which list the authors who made the commits and each commit message (Figure~\ref{fig:typev-deets}).

One issue with analyzing individual authors is that authors can commit from a variety of different computers which can have different emails and usernames attached to the git configuration. This means in the git log one author might appear as several different authors. There is no perfect way to deal with this problem; however, we allow the user of our software to select and view multiple users at a time. Thus allowing the user to view changes from an individual or group of developers.

To allow for a more detailed analyzes of a specific component of the repository we added some basic filtering tools. The first is that a user can filter by type name. The second allows the user to view changes that fall in a specific start and end date. The third allows them to specify the time period of each bar and this includes hour, day, week, and month. Finally, they can choose to see only declarations, types used in declarations and invocations, or invocations.

@@

In the following sections, we present case studies demonstrating the extra insight that the TypeV method makes available.

@@

Apache Bookkeeper deprecated the method AbstractConfiguration.getLedgerManagerType() in release 4.2.0\footnote{\url{https://bookkeeper.apache.org/docs/r4.2.0/apidocs/deprecated-list.html}} and replaced it with AbstractConfiguration.getLedgerManagerFactoryClass(). If we look at getLedgerManagerType() we can see at the time of the start of the deprecation all instances were removed and getLedgerManagerFactoryClass() was added. After the deprecation, we see no more changes to getLedgerManagerType() and only some additions of getLedgerManagerFactoryClass().

@@

The information regarding which types are used in each change is unique to the AST diff method, so it will be useful to see how changes affect the types in a project rather than just the files involved in a diff. We define type coverage as the number of types that have been touched out of the total number of types during either a certain time period or the entire project. File coverage is the same measurement, but using the number of source files instead. To determine if type coverage gives different information than file coverage, we analyzed four popular Java projects: Antlr, Apache Bookkeeper, Apache Curator and Apache Tika. We counted the file coverage and type coverage on a commit-by-commit basis. That is, each datum represents a commit with three dimensions: file coverage and type coverage, and its commit date. Data is sorted in ascending order of commit date. Table~\ref{tab:summary} lists a summary of the data provided. In general, all measures are strongly correlated, both for Pearson (linear) correlation, and Spearman (rank) correlation. As type coverage and file coverage are both cumulative measures, positive correlation is expected. However, even when the distributions are significantly different (as with Antlr and Curator), the correlation is still strongly positive, but not 100\%. This means that when the coverage for types and files of a commit increase, the relationship is not necessarily 1-to-1. This is remarkable in a language like Java, a classes and files are generally in a 1-to-1 relationship.We conclude from this that type coverage yields related, yet different information than file coverage alone.

@@

When managing group projects in an introductory software engineering course, it is very important to be able to see the contributions that students are making to the group repository to judge learning, progress and collaboration. When supervising 4-5 groups of students, it becomes unfeasible to look through all of the commits of each student. Students may take the wrong approach in a certain piece of code which becomes ingrained in the project and very difficult to change. Many groups have problems with group members who do not complete their assigned work or make contributions which end up being meaningless. Issues like this can be difficult to detect and prevent. As a manager, having a tool like TypeV allows one to quickly summarize the work done by each group member, organized by the types they have been working on. This can show when intervention or guidance is required, without the large time commitment of looking through commits.

The purpose of deprecated software features is to give developers time to change projects to follow a new or better standard without breaking the code. Deprecated features can be marked for removal from a project or have some flaw that would make a different feature the better option. Given the time at which a feature was deprecated, seeing the progress of replacement has some important implications. In the case of a security flaw, that time could indicated times at which systems or users were vulnerable. If a feature is marked for removal, a project using a deprecated type has a limited amount of time to make the changes. Removing all instances of deprecated features could also mean a project can be upgraded to use new tools or features that were not possible while the deprecated features were present.  Here we show examples where deprecated classes were removed f

Bug triaging for large open source projects is a very time consuming task, with hundreds  of bugs being reported daily. Each bug must be processed and assigned to a developer that has the experience required to fix the bug. In order to be effective, the triager has to have knowledge of what each developer has worked on as well as the parts of the code may be affected by the bugs. Being able to directly see what types a developer has worked with over time is important information for triaging. If the type causing the bug is known, the developers which have recently used that type can be assigned the bug. The commit messages for the affected types are also visible, providing greater context about how the type has been used. The developer who is assigned the bug can also use type information to locate other affected pieces of code and know which developers have been using the affected types.

@@

Some commands in git present a problem gathering the statistics. Authors can pull other projects into the repository and these additions will count towards the author's changes. These will show up as massive changes by a single author with many types that they did not actually create. Changes of this size are visible in the graph and can be found by inspecting the commit messages.

We only look at the master branch of a repository and ignore other branches; therefore, we might miss some of the work done by an author in another branch. Going through all of the branches in a repository would take a very long time and contain a large amount of duplicate data. Being able to choose which branch to analyze is something that can be added.
